---
title: "Fixing fixation"
author: Pierce Edmiston
output:
  bookdown::pdf_document2:
    toc: no
    keep_tex: yes
    number_sections: no
---
```{r team-structures-config, include=FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
read_chunk("R/exp1.R")
read_chunk("R/exp2.R")
```

# Experiment 1

```{r exp1, include=FALSE}
```

In Experiment 1, we compared the performance of four generations of diachronic
collaboration via vertical transmission---inheriting from someone else---to the
performance of isolated individuals given four separate sessions (Fig.
\@ref(fig:team-structures-selfother)). Participants in the isolated condition
came back to the experiment for four separate sessions, each 25 minutes long.
Sessions were scheduled at least one day apart. At each session after the
first, these participants inherited the recipes for the tools they had created
at the end of the previous session. Thus, in this experiment, both diachronic
and isolated individuals started each session recreating inherited tools before
extending these to discover new tools.

(ref:team-structures-selfother) Problem solving strategies compared in Experiment 1.

```{r team-structures-selfother, fig.width=6, fig.cap="(ref:team-structures-selfother)"}
crotchet::draw_graphviz("team-structures-selfother", package = "totems")
```

## Results {-}

We start by reporting differences in the total number of tools discovered using
either a diachronic or isolated strategy. To foreshadow our results, we found
that, against our predictions, isolated individuals given four sessions were
able to accumulate more tools than diachronic teams given the same total amount
of time. The remaining analyses examine why an isolated strategy was more
effective than a diachronic strategy. We show that isolated individuals were
able to recreate the tools they had discovered previously much more quickly
than diachronic participants. In investigating why an isolated strategy is more
effective, we uncover a benefit to diachronic collaboration in remediating the
costs of fixation suffered by isolated individuals.

### Total tools {-}

The number of innovations discovered in each diachronic generation and isolated
session are shown in Fig. \@ref(fig:innovations-by-generation-selfother). After
a single 25 minute session, there was no difference between diachronic and
isolated participants, `r exp1$gen1_innovations`. This result was expected,
since after a single session, there was no difference between participants
randomly assigned to either strategy. For diachronic participants, each
new generation was able to discover `r exp1$tools_per_diachronic_gen` more
tools than they inherited, `r exp1$tools_per_diachronic_gen_stats`. However,
isolated individuals discovered `r exp1$tools_per_isolated_gen` more tools than
they inherited at each new session, `r exp1$tools_per_isolated_gen_stats`. That
is, an isolated strategy proved more effective than a diachronic strategy over
generations, `r exp1$innovations_by_inheritance_slope`. After four generations
of diachronic collaboration compared to four sessions of isolated labor, an
isolated strategy was more effective, `r exp1$gen4_innovations`.

(ref:innovations-by-generation-selfother) Innovations discovered by diachronic
and isolated strategies. Each thin line is a diachronic team (**left**) or an
isolated individual (**right**). Thick lines with error bars show model
predictions with ±1 standard error.

```{r innovations-by-generation-selfother, fig.width=5, fig.cap="(ref:innovations-by-generation-selfother)"}
innovations_by_generation_plot
```

### Guessing strategy {-}

In order to determine why an isolated strategy was more effective than a
diachronic strategy, we asked whether isolated individuals employed a more
efficient guessing strategy than diachronic participants. As in Experiment 1, we
measured both the proportion of redundant guesses and unique guesses made by
participants in either strategy, corresponding to an inefficient and efficient
guessing strategy, respectively. Our results are shown in Fig.
\@ref(fig:guessing-strategy-by-generation).

Although participants in the isolated strategy discovered more tools after
four sessions than diachronic teams after four generations, they did not engage
in a more efficient guessing strategy. Unlike in Experiment 1, diachronic
participants did not guess a lower proportion of redundant guesses compared to
isolated individuals, `r exp1$prop_redundant_strategy`. For participants in both
strategies, the proportion of redundant guesses decreased over generations and
sessions, `r exp1$prop_redundant_gen`. However, this decrease was not affected
by strategy, `r exp1$prop_redundant_inter`.

For the proportion of unique guesses, there were no differences between
participants in either strategy, `r exp1$prop_unique_strategy`, or
over generations `r exp1$prop_unique_gen`. The change in proportion
of unique guesses over generations was not affected by strategy,
`r exp1$prop_unique_inter`.

(ref:guessing-strategy-by-generation) Proportion of guesses by participants in
diachronic and isolated conditions over generations.

```{r guessing-strategy-by-generation, fig.width=5, fig.cap="(ref:guessing-strategy-by-generation)"}
prop_guess_types_selfother_plot
```

If participants in the isolated strategy did not guess more efficiently or
less redundantly than participants in the diachronic strategy, how then were they
able to accumulate more tools than participants in the diachronic condition? To
answer this question, we investigated the time spent recreating the inherited
tools versus time spent inventing new tools.

### Learning times {-}

Participants who inherited an innovation record from a previous generation or
from a previous session took on average `r exp1$mean_learning_time` minutes of
the 25 minute session (`r exp1$prop_learning_time`%) to recreate the inherited
tools. We refer to this portion of the experiment we refer to as the learning
period.

We found a large difference between diachronic and isolated individuals in the
length of the learning period. Isolated individuals returning for multiple
sessions took significantly less time to recreate the tools they had invented
at the end of their previous session as compared to diachronic participants who
were playing the Totems game for the first time, `r exp1$learning_times_inter`
(Fig. \@ref(fig:diachronic-v-isolated-inheritance)A).

### Guesses for new discoveries {-}

After the learning period ends, participants then had to discover new tools,
not inherited from an ancestor or a previous session. In this section, we
compared the effectiveness of isolated and diachronic participants in
discovering the first new tool, beyond what was inherited.

We found that diachronic participants were able to discover the first new tool
with fewer guesses than isolated participants (Fig.
\@ref(fig:diachronic-v-isolated-inheritance)B). On average, participants in both diachronic
and isolated strategies took `r exp1$first_discovery_mean` guesses to discover
a new tool not inherited from a previous participant or previous session. Diachronic
participants discovered a new tool in `r exp1$first_discovery_beta` fewer guesses
than isolated individuals, `r exp1$first_discovery_stats`. This difference increased
over generations, `r exp1$first_discovery_by_gen_stats`, indicating that an isolated
strategy was becoming more fixated on previous incorrect solutions. In contrast,
the cost of the first new tool among diachronic collaborators was relatively stable
across generations, even as the number of inherited tools increased.

(ref:diachronic-v-isolated-inheritance) Costs and benefits of diachronic
inheritance relative to individual inheritance. **A.** Learning times by
strategy. Learning time refers to the number of minutes it took to recreate the
inherited items. For diachronic participants (**left**), learning time scaled
linearly with generations, but for isolated participants returning for multiple
sessions (**right**), learning times were stable across sessions. **B**. Number
of guesses required to discover the first new tool each session. For diachronic
participants (**left**), the first new tool was stable across generations, but
for isolated participants (**right**) the cost increased with subsequent
sessions.

```{r diachronic-v-isolated-inheritance, fig.height=6, fig.width=5, fig.cap="(ref:diachronic-v-isolated-inheritance)"}
gridExtra::grid.arrange(
  learning_times_plot + ggtitle("A"),
  first_discovery_by_generation_plot + ggtitle("B"),
  ncol = 1
)
```

## Discussion {-}

In Experiment 1, we found that isolated individuals given four sessions were
able to outperform diachronic individuals each given only 25 minutes. However,
an isolated strategy did not result in more efficient problem solving than
a diachronic strategy. Isolated participants did not guess less redundantly, or
generate a higher proportion of unique guesses, than diachronic participants.
Instead, isolated participants, given multiple sessions, were able to more
quickly recreate the tools they had discovered in their previous session, thus
giving them more time to discover new tools. In contrast, for diachronic
participants, recreating the tools inherited from their ancestor took longer
with each generation, leaving less time for new tool discovery.

We obtained some evidence in support of our hypothesis that a diachronic
strategy may promote a more efficient form of problem solving. We found that
isolated individuals were more likely to become fixated on previous, incorrect
solutions. These fixation effects extended over multiple sessions, causing the
performance of some isolated individuals to stagnate, especially for the first
new tool discovered in each session. Diachronic participants, however, were not
impacted by fixation to previous guesses, and were more easily able to move
beyond the solutions they inherited to discover new solutions. These results
reveal a novel benefit to vertical transmission in relieving the pressures that
may slow individual problem solvers down.

# Experiment 2

```{r exp2, include=FALSE}
```

## Results

```{r yoked-by-generation}
yoked_by_generation_plot
```

# Experiment 3

# General Discussion

# References